Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-07-03 manuscript.qmd
update abstract and report sample sizes automatically

author-note:
  disclosures:
    data-sharing: This is version 1 (last modification {{< meta date-modified >}}) of our preprint published at <TODOK>. All materials (reproducible manuscript, analysis code, data) are available in the project's repository on the Open Science Framework (OSF) at <TODO>. A Quarto Manuscripts website is hosted at <TODO>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of this dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current publication.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: This article presents four studies (one qualitative study with n = 8, three quantitative studies with n = 390, n = 921, n = 931) to investigate whether gold-standard items are helpful for questionnaire construction. We constructed three 12-item scales to measure the physical traits of body height, body weight, and age, based on a 2- and a 6-response category response format. In addition, we assessed numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not always exhibit the highest loading on their latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items and gender, an interpretable, systematic residual variance did not always remain. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold standard items indicating that they do not have the same validity. While these results are in line with existing literature, our analyses also showed that items with two response categories were at least as valid as those with six categories, which contradicts past recommendations. When constructing the questionnaire, the items that are intended to measure the latent variable itself should have the highest loading on the latent variable. If this is not the case, the content validity is at least questionable. Implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be empirically compared and most important, hypotheses about items which should represent the latent variable best should be tested.
keywords: [item construction, latent variable modeling, physical scales, gold standard items, response categories]
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

```{r}
#| label: setup
#| include: false

library(papaja)
library(flextable)
library(tidyverse)
library(foreign)
dat_poly <- read.spss("data/Haupterhebung Studie III (Polytom 6PS).sav", 
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)
dat_dich <- read.spss("data/Haupterhebung Studie III (Dichotom 2PS).sav",
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)

dat_desc <- bind_rows(
  dat_poly |> select(Geschlecht, Bildung, Muttersprache, Alter_realW) |> mutate(categories = "6 categories"),
  dat_dich |> select(Geschlecht, Bildung, Muttersprache, Alter_realW) |> mutate(categories = "2 categories")
)

dat_interviews <- read.spss("data/Stichprobe kognitive Interviews.sav", 
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)

dat_pilot <- read.spss("data/Pilottest_Studie_3.sav", 
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)

@@

@@

In psychological research, studies typically develop questionnaires and use latent variable ap-proaches to test whether they measure the intended latent variables [@tenholtScaleConstructionEvaluation2010]. According to the most applied test models (e.g., factor analytic models) especially item discrimination parameters are interpreted as how well items correlate with the latent variable [see @lordStatisticalTheoriesMental1968]. Highly correlated items with the latent variable indicate that these items are more important to measure the latent variable compared to items with lower correlations with the latent variable. Thus, the item discrimination parameter is typically used to assess the quality of an item measuring the latent variable.

>>>>

```{r}
#| label: setup
#| include: false
# load datasets and results objects
# place chunk before the yaml header, so sample sizes can be reported in the abstract
library(papaja)
library(flextable)
library(tidyverse)
library(foreign)
dat_poly <- read.spss("data/Haupterhebung Studie III (Polytom 6PS).sav", 
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)
dat_dich <- read.spss("data/Haupterhebung Studie III (Dichotom 2PS).sav",
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)
dat_desc <- bind_rows(
  dat_poly |> select(Geschlecht, Bildung, Muttersprache, Alter_realW) |> mutate(categories = "6 categories"),
  dat_dich |> select(Geschlecht, Bildung, Muttersprache, Alter_realW) |> mutate(categories = "2 categories"))
dat_interviews <- read.spss("data/Stichprobe kognitive Interviews.sav", 
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)
dat_pilot <- read.spss("data/Pilottest_Studie_3.sav", 
  to.data.frame = TRUE, use.value.labels = FALSE, na.omit = FALSE)

@@

author-note:
  disclosures:
    data-sharing: This is version 1 (last modification {{< meta date-modified >}}) of our preprint published at <TODOK>. All materials (reproducible manuscript, analysis code, data) are available in the project's repository on the Open Science Framework (OSF) at <TODO>. A Quarto Manuscripts website is hosted at <TODO>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of this dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current publication.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on cognitive interviews (n = `r nrow(dat_interviews)`) and a pilot study (n = `r nrow(dat_pilot)`) with an initial item set, three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two (n = `r nrow(dat_dich)`) or six (n = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we argue that items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically compared and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

In psychological research, studies typically develop questionnaires and use latent variable approaches to test whether they measure the intended latent variables [@tenholtScaleConstructionEvaluation2010]. According to the most applied test models (e.g., factor analytic models) especially item discrimination parameters are interpreted as how well items correlate with the latent variable [see @lordStatisticalTheoriesMental1968]. Highly correlated items with the latent variable indicate that these items are more important to measure the latent variable compared to items with lower correlations with the latent variable. Thus, the item discrimination parameter is typically used to assess the quality of an item measuring the latent variable.
